---
title: "UC Davis DataLab Install Guide"
date: "`r Sys.Date()`"
author: "UC Davis DataLab"
output:
  rmdformats::readthedown:
    highlight: kate
    css: datalab_theme.css
bibliography: references.bib
link-citations: TRUE
csl: apa.csl
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}
# git images
git_url <- "../img/mac_finder_1.png"
git_url2 <- "../img/mac_window_2.png"
git_url19 <- "../img/win_git_install_1_browser.png"
git_url20 <- "../img/win_git_install_2_git page.png"
git_url21 <- "../img/win_git_install_3_downloading.png"
git_url22 <- "../img/win_git_install_4_run_open.png"
git_url23 <- "../img/win_git_install_5_allow_changes.png"
git_url24 <- "../img/win_git_install_6_license.png"
git_url25 <- "../img/win_git_install_7_install_location.png"
git_url26 <- "../img/win_git_install_8_default_components.png"
git_url27 <- "../img/win_git_install_9_checked_components.png"
git_url28 <- "../img/win_git_install_10_start_menu.png"
git_url29 <- "../img/win_git_install_11_editor.png"
git_url30 <- "../img/win_git_install_12_initial_branch.png"
git_url31 <- "../img/win_git_install_13_path.png"
git_url32 <- "../img/win_git_install_14_HTTPS_openssl.png"
git_url33 <- "../img/win_git_install_15_line_ending.png"
git_url34 <- "../img/win_git_install_16_terminal_emulator.png"
git_url35 <- "../img/win_git_install_17_git_pull.png"
git_url36 <- "../img/win_git_install_18_credential_helper.png"
git_url37 <- "../img/win_git_install_19_extra_options.png"
git_url38 <- "../img/win_git_install_20_experimental_options.png"
git_url39 <- "../img/win_git_install_21_install_progress.png"
git_url40 <- "../img/win_git_install_22_install_complete.png"
git_url41 <- "../img/win_git_install_23_deslect_and_finish.png"

# OpenRefine images
or_img01 = "../img/win_or_install_1.jpg"
or_img02 = "../img/win_or_install_2.jpg"
or_img03 = "../img/win_or_install_3.jpg"
or_img04 = "../img/win_or_install_4.jpg"
or_img05 = "../img/win_or_install_5.jpg"

# r images
r_img01 = "../img/win_r_install_1_browser.jpg"
r_img02 = "../img/win_r_install_2_mirror.jpg"
r_img03 = "../img/win_r_install_3_type.jpg"
r_img04 = "../img/win_r_install_4_base.jpg"
r_img05 = "../img/win_r_install_5_download.jpg"
r_img06 = "../img/win_r_install_6_installer.jpg"
r_img07 = "../img/win_r_install_7_license.jpg"
r_img08 = "../img/win_r_install_8_location.jpg"
r_img09 = "../img/win_r_install_9_bits.jpg"
r_img10 = "../img/win_r_install_10_startup.jpg"
r_img11 = "../img/win_r_install_11_tasks.jpg"
r_img12 = "../img/win_r_install_12_wait.jpg"
r_img13 = "../img/win_r_install_13_finish.jpg"
r_img14 = "../img/win_r_install_14_rstudio_site.jpg"
r_img15 = "../img/win_r_install_15_version.jpg"
r_img16 = "../img/win_r_install_16_rstudio_download.jpg"
r_img17 = "../img/win_r_install_17_rstudio_installer.jpg"
r_img91 = "../img/win_r_verify_1.jpg"

# R packages
rpkg_img01 = "../img/win_r_pkg_1.jpg"
rpkg_img02 = "../img/win_r_pkg_2.jpg"
rpkg_img03 = "../img/win_r_pkg_3.jpg"
rpkg_img04 = "../img/win_r_pkg_4.jpg"
rpkg_img05 = "../img/win_r_pkg_5.jpg"
rpkg_img06 = "../img/win_r_pkg_6.jpg"

# R Tools
rt_img01 = "../img/win_rt_install_1.jpg"
rt_img02 = "../img/win_rt_install_2.jpg"
rt_img03 = "../img/win_rt_install_3.jpg"
rt_img04 = "../img/win_rt_install_4.jpg"
rt_img05 = "../img/win_rt_install_5.jpg"
rt_img06 = "../img/win_rt_install_6.jpg"
rt_img07 = "../img/win_rt_install_7.jpg"
rt_img08 = "../img/win_rt_install_8.jpg"
rt_img10 = "../img/win_rt_install_10.jpg"
```

# Overview

This toolkit will walk you through installing some of the most common data science tools used at the UC Davis DataLab and beyond. Use the sidebar on the left to navigate to the specific program you would like to install. If you are taking a class or workshop using these tools, please install them prior to attending.

# Anaconda (Python)

## Introduction

## Anaconda on Windows

## Anaconda on Mac

## Verifying your install

## Installation troubleshooting

If you are not able to successfully install Anaconda on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# DBeaver (SQL Database)

## Introduction

## DBeaver on Windows

## DBeaver on Mac

## Verifying your install

## Installation troubleshooting

If you are not able to successfully install DBeaver on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# Git

## Introduction

Git is a ubiquitous software for *version control*. Version control describes a process of storing and organizing multiple versions (or copies) of documents that you create. Approaches to version control range from simple to complex and can involve the use of various human workflows and/or software applications to accomplish the overall goal of storing and managing multiple versions of the same document(s).  

Most people have a folder/directory somewhere on their computer that looks something like this:

![](`r git_url`)

Or perhaps, this:

![](`r git_url2`)

This is a rudimentary form of version control that relies completely on the human workflow of saving multiple versions of a file. This system works minimally well, in that it does provide you with a history of file versions theoretically organized by their time sequence. But this filesystem method provides no information about how the file has changed from version to version, why you might have saved a particular version, or specifically how the various versions are related. This human-managed filesystem approach is more subject to error than software-assisted version control systems. It is not uncommon for users to make mistakes when naming file versions, or to go back and edit files out of sequence. Software-assisted version control systems such as Git were designed to solve this problem.

If you would like more information on how Git works, read more in the [Git Book](https://git-scm.com/book/en/v2){rel="noopener noreferrer" target="_blank"} for free.

## Git on Windows

Follow these step-by-step instructions if you're installing Git on a Windows machine:

First, launch a web browser, the image below shows the Microsoft Edge browser:

![](`r git_url19`) 

Next, navigate to the following Git download URL in your browser [https://git-scm/com/downloads](https://git-scm.com/downloads){rel="noopener noreferrer" target="_blank"}:

![](`r git_url20`) 

Select "Windows" from the Downloads portion of the Git webpage. Git will display the following page and automatically being downloading the correct version of the Git software.  If the download doesn't start automatically, click on the "Click here to download manually link":

![](`r git_url21`) 

When the download is complete, open/Run the downloaded file (will look different in different browsers, but everyone should know how to do this):

![](`r git_url22`) 

A screen will appear asking for permissions for the Git application to make changes to your device.  Click on the **Yes** button:

![](`r git_url23`) 

Click **Next** to accept the user license:

![](`r git_url24`) 

Leave the default "Destination Location" unchanged (usually C:\\Program Files\\Git) and hit **Next**

![](`r git_url25`)

You will see a screen like the one below asking you to "Select Components":

![](`r git_url26`)

Leave all of the default components selected and also check the boxes next to "Additional Icons" and it's sub-item, "On the Desktop". Your completed configurations window should have the following components selected:

```
Additional Icons 
	-> On the Desktop
Windows Explorer integration
	-> Git Bash Here
	-> Git GUI Here
Git LFS (Large File Support)
Associate .git* configuration files with default text editor
Associate .sh files to be run with Bash
```

And should look like this:

![](`r git_url27`)

After verifying that you have the necessary components selected as per above, hit **Next**.

The next screen will ask you to "Select a Start Menu Folder." Keep the default value of Git and hit **Next**:

![](`r git_url28`)

Leave the default "Use Vim (the ubiquitous text editor) as Git's default editor" selected on the "Choosing the default editor used by Git" screen and hit **Next**:

![](`r git_url29`)

On the next screen, leave the default "let Git decide" option selected and hit **Next**:

![](`r git_url30`)

Leave the default "Git from the command line and also from 3rd-party software" selected and hit **Next**:

![](`r git_url31`)

On the next "Choosing HTTPS transport backend" page leave the default "Use the OpenSSL library" option selected and hit **Next**:

![](`r git_url32`)

Leave the default "Checkout Windows-style, commit Unix-style line endings" selected on the next page and hit **Next**:

![](`r git_url33`)

Keep the default "Use MinTTY (the default terminal of MSYS2)" selected on the "Configuring the terminal emulator to use with Git Bash" window and hit **Next**:

![](`r git_url34`)

Keep the default value of "Default (fast-forward or merge)" on the "Choose the default behavior of 'git pull'" page and hit **Next**:

![](`r git_url35`)

Keep the default value of "Git Credential Manager Core" on the "Choose a credential helper" page and hit **Next**:

![](`r git_url36`)

Keep the default values on the "Configuration extra options" page by keeping "Enable file system caching" checked and "Enable symbolic links" unchecked and then hit **Next**:

![](`r git_url37`)

Make sure that no options are checked in the "Configuring experimental options" page and hit **Install**:

![](`r git_url38`)

After you hit this **Install** button as per above, you will see an install progress screen like the one below:

![](`r git_url39`)

When the install is complete, a new, "Completing the Git Setup Wizard" window like the one below will appear:

![](`r git_url40`)

Make sure that all of the options on this window are unchecked as in the image below and then hit the **Finish** button:

![](`r git_url41`)

This will complete your installation process.

*Windows users should verify that when downloading Git for Windows they have also installed Git Bash, which is necessary for working with Git in command line.*

## Git on Mac

If you are installing Git on a Mac, there is no extra configuration. Simply go the git download page at [https://git-scm.com/downloads](https://git-scm.com/downloads){rel="noopener noreferrer" target="_blank"} and choose the latest version for mac, and run the installer package when it is finished downloading. If you get an “unknown developer” warning during the install process, follow the instructions at the beginning of the video at [https://www.youtube.com/watch?v=__kr-Ew5kbE](https://www.youtube.com/watch?v=__kr-Ew5kbE){rel="noopener noreferrer" target="_blank"} to help you work through this problem.

## Verifying your install

**Whether you're installing on Windows or Mac, note that unlike most applications that you've installed before, you will not find a "Git" application in your programs or applications directory once the installation is complete.** As long as you don't get an explicit error message during the installation process, you can assume that the software was successfully installed. Git is a command-line application with which you interact using the command-line, which we'll cover during the interactive session. If you're already familiar with using command line, you can verify your install by opening the terminal (for Windows that will be Git Bash in your programs menu) and type *git --version*. You should then see a response of your installed version (e.g., git version 2.12.2.windows.2, or git version 2.12.2.mac.2), and not the error "command not found."

## Installation troubleshooting

If you are not able to successfully install Git on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# Jupyter Notebooks

## Introduction

## Jupyter on Windows

## Jupyter on Mac

## Verifying your install

## Installation troubleshooting

If you are not able to successfully install Jupyter Notebooks on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# Linux Subsystem for Windows (LSW; a.k.a. Command Line)

## Introduction

## LSW on Windows

## Verifying your install

## Installation troubleshooting

If you are not able to successfully install the Linux Subsystem for Windows on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# OpenRefine

## Introduction

OpenRefine is an open source tool used to clean and pre-process messy data. While most people are familiar with data cleaning in their coding tool of choice (R, Python, Julia, etc.), OpenRefine is designed to provide powerful cleaning capabilities with minimal overhead. One of the most helpful capabilities of OpenRefine is the ability to check for possible duplicates and misspellings of text data using it's text facet tools.

## OpenRefine on Windows

Open your web browser of choice and navigate to the OpenRefine homepage at [https://openrefine.org/](https://openrefine.org/){rel="noopener noreferrer" target="_blank"}. Click on the download button in the left sidebar.

![](`r or_img01`)

On the download page, scroll to the latest version of OpenRefine and select the **Windows kit**. If you are unsure if you have Java installed on your system, choose the **Windows kit with embedded Java** instead.

![](`r or_img02`)

Once the download has completed, open the zip and move the contents to a convenient location on your computer.

![](`r or_img03`)

Open the resulting directory, and double click on the **openrefine.exe** executable.

![](`r or_img04`)

The OpenRefine executable will start a terminal window, and shortly after launch a tab in your default web browser with the OpenRefine interface.
 
![](`r or_img05`)

## OpenRefine on Mac

If you are installing OpenRefine on a Mac, there is no extra configuration. Simply go the [download page for OpenRefine](https://openrefine.org/download.html){rel="noopener noreferrer" target="_blank"} and choose the latest version for mac. Run the installer package when it is finished downloading. If you receive an error regarding the app being from an unidentified developer, please [follow the instructions here](https://support.apple.com/guide/mac-help/open-a-mac-app-from-an-unidentified-developer-mh40616/mac){rel="noopener noreferrer" target="_blank"}.

## Verifying your install

TODO - need data

## Installation troubleshooting

If you are not able to successfully install OpenRefine on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# R/RStudio

## Introduction

["R"](https://www.r-project.org/about.html){rel="noopener noreferrer" target="_blank"} is both a free and open source programming language designed for statistical computing and graphics, and the software for interpreting the code written in the R language. [RStudio](https://rstudio.com/products/rstudio/){rel="noopener noreferrer" target="_blank"} is an integrative development environment (IDE) within which you can write and execute code, and interact with the R software. It's an interface for working with the R software that allows you to see your code, plots, variables, etc. all on one screen. This functionality can help you work with R, connect it with other tools, and manage your workspace and projects. You cannot run RStudio without having R installed. While RStudio is a commercial product, the free version is sufficient for most researchers.

**Why learn R?** There are many advantages to working with R. 

* **Scientific integrity.** Working with a scripting language like R facilitates reproducible research. Having the commands for an analysis captured in code promotes transparency and reproducibility. Someone using your code and data should be able to exactly reproduce your analyses. An increasing number of research journals not only encourage, but are beginning to require, submission of code along with a manuscript. 
* **Many data types and sizes.** R was designed for statistical computing and thus incorporates many data structures and types to facilitate analyses. It can also connect to local and cloud databases.
* **Graphics.** R has built-in plotting functionalities that allow you to adjust any aspect of your graph to effectively tell the story of your data.
* **Open and cross-platform.** Because R is free, open-source software that works across many different operating systems, anyone can inspect the source code, and report and fix bugs. It is supported by a large community of users and developers.
* **Interdisciplinary and extensible.** Because anyone can write and share R packages, it provides a framework for integrating approaches across domains, encouraging innovation.

## R/RStudio on Windows

Follow these step-by-step instructions to install R and RStudio on a Windows machine:

First open your internet browser of choice, and navigate to [https://www.r-project.org/](https://www.r-project.org/){rel="noopener noreferrer" target="_blank"}. Click on **download R**.

![](`r r_img01`)

On the following page, select the link under whatever location is closest to you for the best download speed (though any will work).

![](`r r_img02`)

Next, click the **Download R for Windows** link.

![](`r r_img03`)

Click on the link **base** to go to the download page.

![](`r r_img04`)

Finally, click **Download R X.X.X for Windows** to download the installer.

![](`r r_img05`)

When the download is complete, run the R installer. This will look slightly different depending on your browser.

![](`r r_img06`)

Select your language and then accept the license agreement by hitting **Next >**.

![](`r r_img07`)

Leave the default install location and select **Next >**.

![](`r r_img08`)

If you know what kind of machine you are on, you can specify if you want the 32 or 64 bit version of R. If you do not know, it is safe to install both.

![](`r r_img09`)

Keep the default startup options and hit **Next >**.

![](`r r_img10`)

You most likely will not want an R shortcut on your desktop, as you will almost certainly use RStudio as an interface. You can still have one if you would like. Otherwise, accept the defaults and hit **Next >**.

![](`r r_img11`)

Wait for the instillation to complete.

![](`r r_img12`)

Once it is done, hit **Finish**. You've now installed R! However, we still need to install RStudio separately.

![](`r r_img13`)

Navigate to the RStudio homepage at [https://rstudio.com/](https://rstudio.com/){rel="noopener noreferrer" target="_blank"} and click the download button.

![](`r r_img14`)

Scroll down and select the free version to download. If you are using RStudio for commercial purposes you will need to look into RStudio's licensing terms to see if you need to pay for the pro version.

![](`r r_img15`)

Download the RStudio installer for your machine.

![](`r r_img16`)

Run the installer just as you did for the R download.

![](`r r_img17`)

## R/RStudio on Mac

If you are installing R/RStudio on a Mac, there is no extra configuration. Simply go the download pages for [R](https://www.r-project.org/){rel="noopener noreferrer" target="_blank"} and [RStudio](https://rstudio.com/){rel="noopener noreferrer" target="_blank"} and choose the latest version for mac. Run the installer package when it is finished downloading. If you receive an error regarding the app being from an unidentified developer, please [follow the instructions here](https://support.apple.com/guide/mac-help/open-a-mac-app-from-an-unidentified-developer-mh40616/mac){rel="noopener noreferrer" target="_blank"}.

## Verifying your install

Once you have installed **both** R and RStudio, you should be able to run RStudio on your machine. You can verify your install is working by opening RStudio and typing `paste("Hello World!")` into the console as shown below. If the code runs you should see a response that says `[1] Hello World!`. If that works you are all set!

![](`r r_img91`)

## Installation troubleshooting

If you are not able to successfully install R/RStudio on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# R Packages

## Introduction

R packages are community made functions that automate or expand the things you can do in the R language. The process for installing them is largely the same for both Windows and Mac. There are three main methods for installing R packages. You can install them from the Comprehensive R Archive Network (CRAN) from within R, from another online source like [Github](https://github.com/){rel="noopener noreferrer" target="_blank"}, or from files on your local machine. We will cover each of these methods here.

## Packages from CRAN

The vast majority to packages in R can be installed from CRAN. You will need the name of the package you want to install. Once you have the name (case sensitive!), you can download it from within R using the `install.packages()` function in the R console.

For example, if you wanted to install the `skimr` package, you would enter `install.packages("skimr")`. Note that you do need to put the package name in quotes, and that it is case sensitive. You can install multiple packages at a time by passing a vector of package names to `install.packages()`, for example: `install.packages(c("skimr", "corrplot"))`.

You can test if the packages installed correctly by calling it in R using the `library()` function. 

## Pacakges from Github

Not all packages are available on CRAN, especially very new or very old packages. For these you will most likely need to install them from an online repository, the most common of which is [Github](https://github.com/){rel="noopener noreferrer" target="_blank"}. There are some R packages that make this process easy. We will use the `remotes` package.

First install the `remotes` package by running `install.packages("remotes")` in the R console. If it installed correctly, you will see `package ‘remotes’ successfully unpacked and MD5 sums checked`.

![](`r rpkg_img01`)

Once `remotes` is installed, you can install packages directly from an online repository. We will use `remotes`'s `install_github()` function to demonstrate. Let's try installing the handy [`wordcountaddin`](https://github.com/benmarwick/wordcountaddin){rel="noopener noreferrer" target="_blank"} for RStudio. First navigate to the package page on Github: [https://github.com/benmarwick/wordcountaddin](https://github.com/benmarwick/wordcountaddin){rel="noopener noreferrer" target="_blank"}.

![](`r rpkg_img02`)

Most packages that need to be installed from an online source will give you the code to do so in the readme of their repository. Another common package for installing from online repositories is called `devtools`, but this package contains many other functions we don't need for this task. You can replace `devtools` with `remotes` almost all of the time.

![](`r rpkg_img03`)

For our case, type `remotes::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)` into the console and hit enter. You may be asked if you would like to update other packages. Enter `1` to try and update all. If this doesn't work, you can enter the command again and try `3` to ignore these updates.

![](`r rpkg_img04`)

Read the resulting install output. As long as it does not indicate anything failed to install, you're done!

![](`r rpkg_img05`)

You can now use the helpful `wordcountaddin`! You can find it in the **Addins** dropdown above the editor pane.

![](`r rpkg_img06`)

## Packages from Source

# R Tools

## Introduction

R Tools is a bundle of programs on Windows that allows R to build packages from local source files, rather than installing through CRAN. The vast majority of the time this is unnecessary, but some circumstances require it. Mac users do not need to install R Tools. 
## RTools on Windows

First, navigate to the R tools website at [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/){rel="noopener noreferrer" target="_blank"}.

![](`r rt_img01`)

Scroll down to the download links and select the **rtoolsXX-x86_64.exe** link.

![](`r rt_img02`)

Once the download has finished, run the R tools installer.

![](`r rt_img03`)

On the first page of the installer, select a custom location if you would like, but otherwise press **Next >**.

![](`r rt_img04`)

Leave the defaults for the additional tasks and hit **Next >**.

![](`r rt_img05`)

On the next page, hit the install button and wait for it to complete.

![](`r rt_img06`)

One the install is done, press **Finish**. You're not done yet though!

![](`r rt_img07`)

In order for R to make use of R tools, you need to add it to the `PATH` that R looks for tools on. To do this, open R or R Studio and type `writeLines('PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"', con = "~/.Renviron")` in the console and press enter.

![](`r rt_img08`)

## Verifying your install

To verify R tools was installed successfully, first restart R to assure you are in a clean environment. You can do this by closing and re-opening your R or RStudio window. Afterwards, type `Sys.which("make")` (case sensitive!) into the console. If you see `"C:\\rtools40\\usr\\bin\\make.exe"` as a result you are all good!

![](`r rt_img10`)

## Installation troubleshooting

If you are not able to successfully install R Tools on your own, please attend DataLab's Virtual Office Hours. Click [here](https://datalab.ucdavis.edu/office-hours/){rel="noopener noreferrer" target="_blank"} for more information and to receive a Zoom link.

# Contributions

This research toolkit is maintained by the UC Davis DataLab, and is open for contribution. See how you can contribute on the [Github repo](https://github.com/ucdavisdatalab/research-toolkits){rel="noopener noreferrer" target="_blank"}.

This toolkit has been made possible thanks to contributions by:

* Carl Stahmer
* [Jared Joseph](https://jnjoseph.com/){rel="noopener noreferrer" target="_blank"}